---
title: 'STAT636: Partial Correlation'
subtitle: Let's investigate partial correlation
output:
  html_document: default
---


# Summary
This document investigates using the bootstrap to make confidence intervals for partial correlations and compares them to correlation

# Running example: Simulation

Here, we are going to generate random vectors that have a specified precision matrix Omega.  Here, the precision matrix has a block diagonal
structure.  Later, we will change rho2 to break this block diagonal structure
```{r simulate1}
n           = 400
p           = 5

#adding partial correlation
rho         = .1
rho2        = .0
Omega       = matrix(c(1,rho,rho, rho2,0,
                       rho, 1 , rho, 0,0,
                       rho, rho, 1 , 0, 0,
                       rho2, 0, 0 , 1 , rho,
                       0, 0,  0, rho, 1),nrow=p)

Omega
Sigma      = solve(Omega)
Sigma
R          = -diag(1/sqrt(diag(Omega))) %*% Omega  %*% diag(1/sqrt(diag(Omega)))
set.seed(1)
X          = matrix(rnorm(n*p),nrow=n) 
Xpc        = X %*% chol(Sigma)
```

#Computing the estimators

We can compute the estimators as:
```{r estimator1}
Spc         = cov(Xpc)
OmegaHat    = solve(Spc)
Rhat        = -diag(1/sqrt(diag(OmegaHat))) %*% OmegaHat  %*% diag(1/sqrt(diag(OmegaHat)))
```

# The bootstrap

Let's get some bootstrap confidence intervals
```{r bootstrap1}
B = 10000
Rhat_starOutput = array(dim=c(p,p,B))
for(b in 1:B){
	bootSamp      = sample(1:n,n,replace=TRUE)
	Xpc_star      = Xpc[bootSamp,]
	Spc_star      = cov(Xpc_star)
	OmegaHat_star = solve(Spc_star)
	Rhat_star     = -diag(1/sqrt(diag(OmegaHat_star))) %*% OmegaHat_star  %*% diag(1/sqrt(diag(OmegaHat_star)))
	Rhat_starOutput[,,b] = Rhat_star	
}
apply(Rhat_starOutput,1:2, quantile, 0.025)
apply(Rhat_starOutput,1:2, quantile, 0.975)
```

Let's check and see if the true parameter is contained in the bootstrap intervals
```{r checkIntervals1}
R > apply(Rhat_starOutput,1:2, quantile, 0.025)
R < apply(Rhat_starOutput,1:2, quantile, 0.975)
```

# Graphical models

Let's plot the partial correlation graphical model.  First, here is the truth:

```{r partialCorrelation1}
require(igraph)
edge = abs(R) > 0
g    = graph.adjacency(edge , mode="undirected", diag=FALSE )
plot(g, layout=layout.fruchterman.reingold)
```

We want to add an edge if 0 is not in the associated bootstrap interval
```{r partialCorrelationHat1}
noEdge = 0 > apply(Rhat_starOutput,1:2, quantile, 0.025) & 0 < apply(Rhat_starOutput,1:2, quantile, 0.975)
g = graph.adjacency(!noEdge , mode="undirected", diag=FALSE )
plot(g, layout=layout.fruchterman.reingold)
```

Let's compare the correlation graphical model.  Here is the theoretical graph:
```{r correlation1}
edge = abs(Sigma) > 0
g = graph.adjacency(edge , mode="undirected", diag=FALSE )
plot(g, layout=layout.fruchterman.reingold)
```

And here is the estimated one, without multiple correction:
```{r correlationHat1}
alpha = 0.05
fisherStat = 1/2 * log((1+cor(Xpc))/(1-cor(Xpc)))
diag(fisherStat) = 0

#CI
untransformedLower = fisherStat - qnorm(1 - alpha/2)/sqrt(n-3)
untransformedUpper = fisherStat + qnorm(1 - alpha/2)/sqrt(n-3)
transformedLower   = (exp(2*untransformedLower)-1)/(exp(2*untransformedLower)+1)
transformedUpper   = (exp(2*untransformedUpper)-1)/(exp(2*untransformedUpper)+1)

noEdge = 0 > transformedLower & 0 < transformedUpper
g = graph.adjacency(!noEdge , mode="undirected", diag=FALSE )
plot(g, layout=layout.fruchterman.reingold)
```

Now with multiple correction:
```{r correlationHatCorrection1}
alpha = 0.05/10
#CI
untransformedLower = fisherStat - qnorm(1 - alpha/2)/sqrt(n-3)
untransformedUpper = fisherStat + qnorm(1 - alpha/2)/sqrt(n-3)
transformedLower   = (exp(2*untransformedLower)-1)/(exp(2*untransformedLower)+1)
transformedUpper   = (exp(2*untransformedUpper)-1)/(exp(2*untransformedUpper)+1)

noEdge = 0 > transformedLower & 0 < transformedUpper
g = graph.adjacency(!noEdge , mode="undirected", diag=FALSE )
plot(g, layout=layout.fruchterman.reingold)
```
